---
title: "Interactions"
format: 
  html:
    fig-width: 10
execute:
  message: false
  warning: false
bibliography: references.bib
---

::: callout-warning
THIS IS A DRAFT.
:::

## Set up

```{r}
library(tidyverse)
library(polisciols)
library(modelsummary)
library(marginaleffects)
library(broom)
```

## Exporting mass destruction

Trade between states is not only the product of economic considerations. States often promote or restrict trade with other states for political and security reasons. International political economists have identified a wide range of political factors that shape international trade, including regime type (democracies tend to trade more with each other than with other regimes), instability (states tend to trade less with states they believe to be at a high risk of conflict), and alliances (trade tends to "follow the flag").   

These theories of the political determinants of trade flows are often agnostic to the good being traded. But this is not necessarily the case. It is reasonable to expect that a state will be more mindful of who it supplies javelin missiles to than to whom it sells typewriters. @fuhrmann2008 explores this distinction explicitly in his paper *Exporting Mass Destruction? The Determinants of Dual-Use Trade*. 

He seeks to identify the strategic determinants of trade in dual-use commodities. These are commodities that can be used to produce Weapons of Mass Destruction (WMDs) or in legitimate civilian applications. Unlike javelin missiles, the purpose to which importing states will put these commodities is not clear. How do exporters balance the benefits of trade against the risk that they may enable WMD manufacturing? 

To explore this question, Fuhrmann analyzes exports of dual-use commodities from the US to all other countries in the post-Cold War period between 1991 and 2001. Whilst he explores both economic and political determinants, his focus is on whether or not the US and its trade partner share the same foreign policy goals. 

Before we dive into his full model, let's set up our expectations. We should expect that states will be more reticent to sell these sensitive commodities to their enemies. Who knows to what nefarious purpose they may put them to? 

Fuhrmann uses the degree to which states share similar foreign policy platforms to identify friends and enemies. This variable, `pol_dissimiliarity` is binary and takes on a value of `1` if the country enacts foreign policies that are very dissimilar to those of the US. In other words, `pol_dissimilarity = 1` when they are enemies.      

::: {.callout-note}
For those who are interested, he asks whether the average compatibility between the country's and the United States' foreign policy between 1991 and 2001 was in the lowest 10th percentile of all countries' similarities with the United States. 
:::

So, we should expect that the US traded less in these sensitive commodities with its enemies than with its friends. Let's take a look: 

```{r}
x_mass_destruction |> 
  group_by(pol_dissimilarity) |> 
  summarise(avg_x_dual_use = scales::dollar(mean(x_dual_use)))
```

Hmm, the US appears to have provided its enemies with more dual-use commodities than its friends! That's not very strategic! 

But wait, let's not write off US foreign policy just yet. Dual-use commodities are sensitive because they can be used to produce or support WMDs. But not all countries that import these commodities pursue WMDs. Many need these commodities for civilian applications. 

So, it may in fact be the case that the US is only reticent to sell dual-use commodities to enemies that it knows are pursuing or have access to WMDs. Let's include this characteristic in our summary:

```{r}
x_mass_destruction |> 
  group_by(pol_dissimilarity, wmd_pursuit) |> 
  summarise(avg_x_dual_use = scales::dollar(mean(x_dual_use))) 
```

Ah! There we go! The US is happy to trade in dual-use commodities with enemies when they are not pursuing WMDs. Further, the US is happy to trade in dual-use commodities with its friends that *are* pursuing WMDs. I guess their interests must be really aligned.   

This simple summary suggests that the relationship between trade in dual-use commodities and foreign policy is not as straight forward as we first thought. The US appears to be less willing to trade in these sensitive commodities with enemies *if they are pursuing WMDs*. This condition is critical to understanding this relationship. 

In other words, we cannot say anything about the relationship between trade in dual-use commodities and foreign policy without first knowing whether the country in question is pursuing WMDs. If someone were to ask you "do you think the US avoids exporting dual-use commodities to its enemies?", you should reply "it depends!".      

We can explore this more complex relationship formally using **interactions**. Below, we will replicate Fuhrmann's analysis. In doing do, we will strengthen our understanding of multiple linear regression and introduce the knowledge required to interpret interaction effects. 

Let's get started!

## Dual-use trade

Fuhrmann draws on existing literature to build an understanding of the determinants of trade in dual-use commodities. Let's build his baseline model of US exports of dual-use commodities: 

```{r}
m <- lm(log(x_dual_use) ~ defence_ally + democracy + conflict + pol_dissimilarity + 
          wmd_pursuit + wmd_acquisition + log_gdp + log_population + 
          log_distance + log_x_all,
        data = x_mass_destruction)

modelsummary(m, 
             stars = T,
             coef_rename = c(
               "log_gdp" = "Average GDP (USD logged)",
               "log_population" = "Average population (logged)",
               "log_distance" = "Distance between the United States and the country (logged)",
               "log_x_all" = "Total imports from the US (USD logged)"))
```

His argument focuses the level of political similarity between the US and its trade partner. Let's look at the value of dual-use exports this model predicts friends and foes to receive:  

```{r}
plot_predictions(m, condition = "pol_dissimilarity")
```

As above, we see that countries with strongly dissimilar foreign policy agendas to the US receive the most dual-use exports from it. This runs counter to our expectations. 

This model assumes that the association between exports in dual-use commodities and the level of compatibility between the trade partner's and the United States' foreign policy is constant across all trade partners regardless of their desire or capacity to produce WMDs. However, as Fuhrmann argues, this should not be the case: the US should be much more reticent to export dual-use commodities to states that are pursuing foreign policies that are not aligned with the US's interests *and* seek to produce WMDs. 

We can capture this by introducing into our model an interaction between dissimilarities and whether a state is pursuing acquiring nuclear, chemical, or biological weapons: 

```{r}
m <- lm(log(x_dual_use) ~ defence_ally + democracy + conflict + pol_dissimilarity + 
          wmd_pursuit + wmd_acquisition + wmd_pursuit * pol_dissimilarity + log_gdp + 
          log_population + log_distance + log_x_all,
        data = x_mass_destruction)

modelsummary(m)
```

::: {.callout-tip}
To include an interaction in your `lm()` function, multiply the two variables together. Above, I included an interaction between `pol_dissimilarity` and `wmd_pursuit` by including *in addition to the two variables on their own* `pol_dissimilarity * wmd_pursuit`.

You must also include the interacted variables on their own. Above I included: 

```{r}
#| eval: false

lm(log(x_dual_use) ~ ... + pol_dissimilarity + wmd_pursuit + wmd_pursuit * pol_dissimilarity, 
   data = x_mass_destruction)
```
:::

You need to be very careful when interpreting these coefficients. Let's go through some of the important ones together. 

One way to think about interactions is to look at the different regression lines for trade partners that are politically dissimilar and for those that are similar to the US. 

Let's start with the general regression function:

$$
Dual\ use\ exports = −2.82 + ... + 2.16 * pol\_dissimilarity + 2.20 * wmd\_pursuit - 4.20 pol\_dissimilarity * wmd\_pursuit + \epsilon
$$

Now, let's look at this function for states that share the US's foreign policy goals (`pol_dissimilarity = 0`):

$$
Dual\ use\ exports = −2.82 + ... + 2.16 * 0 + 2.20 * wmd\_pursuit - 4.20 * 0 * wmd\_pursuit + \epsilon
$$

$$
Dual\ use\ exports = −2.82 + ... + 2.20 * wmd\_pursuit + \epsilon
$$

And for states that are dissimilar to the US: 

$$
Dual\ use\ exports = −2.82 + ... + 2.16 * 1 + 2.20 * wmd\_pursuit - 4.20 * 1 * wmd\_pursuit + \epsilon
$$

$$
Dual\ use\ exports = -0.66 - 2 * wmd\_pursuit + \epsilon
$$

So, among states that share the US's political goals, pursuing WMDs actually increases their dual-use exports from the US relative to those like-minded states that do not pursue WMDs. We predict that the value of dual-use exports for these countries is -0.66 logged dollars, compared to −2.82 logged dollars among like-minded states that are not pursuing these weapons. 

We can contrast this with our expectations among states that do not share the US's political goals. When these states are pursuing WMDs, they receive -2.66 logged dollars of exports in these useful commodities from the US. As expected, this is less than that received by dissimilar states that are not pursuing these weapons (who we predict will receive -0.66 logged dollars in exports). 

Let's make this a little more tangible. Let's find the average value of dual-use exports our model predicts friends and foes pursuring and not pursuing WMDs to receive: 

```{r}
augment(m) |> 
  group_by(pol_dissimilarity, wmd_pursuit) |> 
  summarise(pred_x_dual_use = scales::dollar(exp(mean(.fitted))))
```

As expected, friends who pursue WMDs are predicted to receive more dual-use exports than friends who are not pursuing these weapons. Foes who pursue WMDs are predicted to receive less dual-use exports than foes who are not pursuing these weapons.

More generally, the coefficient of political dissimilarity represents *the difference between* the predicted value of dual-use exports (logged) between the US and states that are politically similar to the US and that did not pursue acquiring WMDs. 

The coefficient of WMD pursuit represents the *difference between* the predicted value of dual-use exports (logged) between the US and states that are not pursuing acquiring WMDs and those that are. 

The coefficient of the interaction term compares the predicted value of dual-use exports (logged) between the US and states that did or did not pursue acquiring WMDs. 
