---
title: "Consistency and bias"
subtitle: "GVPT722"
format: 
  revealjs:
    slide-number: true
    preview-links: auto
    theme: solarized
    embed-resources: true
execute: 
  message: false
  warning: false
  echo: true
---

## What problem are we solving?

*Building confidence in your inference from a finite number of random samples.*

## Data and packages

```{r}
library(tidyverse)
library(poliscidata)
library(modelsummary)
library(broom)
library(ggdist)
```

Because we are working with randomness:

```{r}
set.seed(222)
```

## Consistency

Refers to the probability that each random sample from our population will produce a similar set of estimates of our regression coefficients.

## Consistency

*What is the relationship between an individual's feelings towards President Obama and their party affiliation?*

```{r}
poliscidata::nes |> 
  select(caseid, obama_therm, dem) |> 
  glimpse()
```

## Consistency

*What is the relationship between an individual's feelings towards President Obama and their party affiliation?*

```{r}
m <- lm(obama_therm ~ dem, data = nes)
```

```{r}
#| echo: false

modelsummary(m, 
             coef_rename = c(dem = "Democrat"), 
             statistic = NULL, 
             stars = T,
             gof_map = c("nobs"))
```

::: aside
Individuals who identify as Democrats have, on average, `r tidy(m) |> filter(term == "dem") |> pull(estimate) |> round(1)` points warmer feelings towards President Obama than individuals who do not identify as Democrats. 
:::

## Random sample from our "population"

Let's start with a random sample of 100 respondents:

```{r}
nes_100 <- nes |> 
  sample_n(100) |> 
  select(caseid, obama_therm, dem)

glimpse(nes_100)
```

## Learning from this sample

```{r}
m <- lm(obama_therm ~ dem, data = nes_100)
```

```{r}
#| echo: false

modelsummary(m, 
             coef_rename = c(dem = "Democrat"), 
             statistic = NULL, 
             stars = T,
             gof_map = c("nobs"))
```

## Acknowledging randomness

Let's take a different random sample of 100 respondents:

```{r}
nes_100 <- nes |> 
  sample_n(100) |> 
  select(caseid, obama_therm, dem)

glimpse(nes_100)
```

## Acknowledging randomness

Let's take a different random sample of 100 respondents:

```{r}
m <- lm(obama_therm ~ dem, data = nes_100)
```

```{r}
#| echo: false

modelsummary(m, 
             coef_rename = c(dem = "Democrat"), 
             statistic = NULL, 
             stars = T,
             gof_map = c("nobs"))
```

## Acknowledging randomness

Let's take 1,000 different random samples of 100 respondents. 

```{r}
#| echo: false

random_sample_nes <- function(n) {
  
  nes_sample <- nes |> 
    sample_n(n) |> 
    select(caseid, obama_therm, dem)
  
  m <- lm(obama_therm ~ dem, data = nes_sample)
  
  dem_estimate <- tidy(m) |> 
    filter(term == "dem") |> 
    pull(estimate)
  
  dem_estimate
  
}

nes_100_dem_estimates <- map(rep(100, 1000), random_sample_nes)

ggplot(tibble(beta_dem = unlist(nes_100_dem_estimates)), 
       aes(x = beta_dem)) + 
  stat_halfeye() + 
  theme_minimal()
```

## Building confidence in our one random sample

We can only take a finite number of random samples from our population.

-   How can we increase our confidence that the estimated coefficients produced by these random samples is close to the truth? 

## Building confidence in our one random sample

We can do this by increasing our sample size. The larger the sample size, the more consistent the estimates.

-   Let's look at 1,000 different random samples of 300 respondents. 

-   And 1,000 different random samples of 1,000 respondents. 

Do we get more consistent estimates? 

## Building confidence in our one random sample

```{r}
#| echo: false

nes_300_dem_estimates <- map(rep(300, 1000), random_sample_nes)

nes_1000_dem_estimates <- map(rep(1000, 1000), random_sample_nes)

tibble(sample_size = 100, 
       beta_dem = unlist(nes_100_dem_estimates)) |> 
  bind_rows(
    tibble(
      sample_size = 300,
      beta_dem = unlist(nes_300_dem_estimates)
    )
  ) |> 
  bind_rows(
    tibble(
      sample_size = 1000,
      beta_dem = unlist(nes_1000_dem_estimates)
    )
  ) |> 
  ggplot(aes(x = beta_dem, fill = factor(sample_size))) + 
  stat_halfeye(alpha = 0.5) + 
  theme_minimal() + 
  theme(legend.position = "bottom") + 
  labs(fill = "Sample size")
```

## Bias

*A biased coefficient estimate will systematically be higher or lower than the true value.*

## What if we only sampled from males? 

What happens to our understanding of the relationship between an individual’s feelings towards Obama and their party affiliation?

```{r}
nes_men <- nes |> 
  filter(gender == "Male") |> 
  select(caseid, obama_therm, dem, gender)

glimpse(nes_men)
```

## A consistent but biased estimate

Let's take a random sample of 1,000 individuals from this male-only pool.  

```{r}
#| echo: false

nes_1000_men <- sample_n(nes_men, 1000)

m_men <- lm(obama_therm ~ dem, data = nes_1000_men)

modelsummary(list("Males only" = m_men), 
             coef_rename = c(dem = "Democrat"), 
             statistic = NULL, 
             stars = T,
             gof_map = "nobs")
```

## A consistent but biased estimate

```{r}
#| echo: false

nes_1000 <- sample_n(nes, 1000)

m_all_sample <- lm(obama_therm ~ dem, data = nes_1000)
m_all <- lm(obama_therm ~ dem, data = nes)

modelsummary(list("Males only" = m_men, 
                  "All respondents" = m_all_sample,
                  "'True' relationship" = m_all), 
             coef_rename = c(dem = "Democrat"), 
             statistic = NULL, 
             stars = T,
             gof_map = "nobs")
```

## Summary

-   Cannot take infinite samples from our population

-   Can use our understanding of uncertainty to increase our confidence in a single or finite number of random samples from our population (*consistency*)

-   Need to ensure that we are not excluding groups of observations from the population from which we draw those random samples (*bias*)

-   **We aim to have consistent *and* unbiased estimates of our coefficients**
